Translation of B Implementations to the LLVM: Control and Modularity

### 2013-09-06 translation.tex
Updated examples so that they are built with the latest version of bimp.py Added an example to test translation of an implementation with an import clause without prefix. Implemented translation of import clauses, and call instructions. Updated the specification of the translation.

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\text{\Return}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

\paragraph{Becomes equal instructions} must evaluate the target and the source
of the assignment, and then copy the result of the latter in the former, using
the LLVM instruction \llvm{store}. The evaluation of the target is defined with
$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
variable $p$ containing the assigned location. The evaluation of the source
yields the corresponding code $r$, the variable $v$ holding the result, and its
type $t$.
\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef 
  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
%   & l \conc r \conc 
%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
%   \conc t \conc \llvm{"* "} \conc p \conc \nl
\end{split}
\end{align*}
The assignment target may be either local to the operation or a state
variable. In the former case, it can be referenced by the corresponding LLVM
identifier. In the latter case, it is represented as an element of the state
structure $\Sigma$, and its address must be calculated and assigned to a new temporary
variable.
\begin{align*}
\lefteqn{\trad{n}{lv} \isdef} \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""}, \local(\B{n}) \\
& \quad \ELSE \LET v = \variable \AND t = \trad{n.type}{\B{Type}} \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v
\end{align*}

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generator produces instructions to evaluate the operation
arguments, the address of the variables receiving the results as well as an
instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated (in the case of a so-called local
operation), or an instance of an imported machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \Global(\B{c}) \AND ty = \state(\B{c.root}) \AND \\
  & \quad val = \IF \B{c.inst} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})}"}\\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
\end{align*}

@@

For each test file, indicates the B0 constructs found in the implementation.
\begin{center}
\newcommand{\no}[0]{--}
\newcommand{\yes}[0]{$\circ$}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Example nº & 01 & 02 & 03 & 04 & 05 & 06 \\
\hline
\hline
\multicolumn{7}{|c|}{clauses} \\
\hline
variables & \yes & \yes & \yes & \yes & \yes & \yes\\
constants & \no & \yes & \no & \no & \yes & \no \\
op. input & \yes & \no & \yes & \no & \no & \no \\
op. output & \yes & \no & \yes & \no & \no & \no \\
\hline
\multicolumn{7}{|c|}{instructions} \\
\hline
block & \yes & \no & \yes & \yes & \no & \yes\\
local var & \no & \no & \yes & \yes & \no & \yes \\
identity & \no & \no & \yes & \no & \no & \no \\
becomes equal & \yes & \yes & \yes & \yes & \yes & \yes\\
operation call & \no & \no & \no & \no & \no & \no \\
condition & \yes & \no & \no & \no & \yes & \no \\
case & \no & \no & \no & \no & \no & \yes \\
assertion & \no & \no & \no & \no & \no & \no \\
sequence & \yes & \yes & \yes & \yes & \no & \yes \\
while & \no & \no & \no & \yes & \no & \no\\
\hline
\multicolumn{7}{|c|}{types} \\
\hline
INT & \yes & \yes & \yes & \yes & \yes & \yes \\
BOOL & \yes & \yes & \no & \no & \yes & \no \\
\hline
\multicolumn{7}{|c|}{expressions} \\
\hline
int lit & \yes & \yes & \yes & \yes & \yes & \yes\\
bool lit & \yes & \yes & \no & \yes & \yes & \no \\
vars & \yes & \yes & \yes & \yes & \no & \yes \\
loc. vars & \no & \no & \yes & \yes & \no & \yes \\
const & \no & \yes & \no & \no & \yes & \no \\ 
succ & \no & \yes & \no & \yes & \yes & \no \\
pred & \no & \yes & \no & \yes & \yes & \no \\
+ & \yes & \no & \no & \yes & \no & \yes \\
- & \yes & \no & \no & \no & \no & \no \\
* & \no & \no & \no & \no & \no & \yes \\
\hline
\multicolumn{7}{|c|}{formulas} \\
\hline
and & \no & \no & \no & \no & \yes & \no \\
or & \no & \no & \no & \no & \yes & \no \\
not & \no & \no & \no & \no & \yes & \no \\
$<$ & \yes & \yes & \no & \no & \no & \no \\
$>$ & \no & \no & \no & \no & \yes & \no \\
\hline
\end{tabular}
\end{center}

>>>>

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

\paragraph{Becomes equal instructions} must evaluate the target and the source
of the assignment, and then copy the result of the latter in the former, using
the LLVM instruction \llvm{store}. The evaluation of the target is defined with
$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
variable $p$ containing the assigned location. The evaluation of the source
yields the corresponding code $r$, the variable $v$ holding the result, and its
type $t$.
\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef 
  & \LET l \sep p \sep t'= \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
%   & l \conc r \conc 
%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
%   \conc t \conc \llvm{"* "} \conc p \conc \nl
\end{split}
\end{align*}
The assignment target may be either local to the operation or a state
variable. In the former case, it can be referenced by the corresponding LLVM
identifier. In the latter case, it is represented as an element of the state
structure $\Sigma$, and its address must be calculated and assigned to a new temporary
variable.
\begin{align*}
\lefteqn{\trad{n}{lv} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generator produces instructions to evaluate the operation
arguments, the address of the variables receiving the results as well as an
instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated (in the case of a so-called local
operation), or an instance of an imported machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \Global(\B{c}) \AND ty = \llvm{"\PH{\state(\B{c.root})} *} \AND \\
  & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
  & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
  & \quad \quad \ELSE \LET l = \lbl \IN \\
  & \quad \quad \quad l \sep \\
  & \quad \quad \quad \llvm{"\PH{l} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{\PH{p}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
\end{align*}

@@

For each test file, indicates the B0 constructs found in the implementation.
\begin{center}
\newcommand{\no}[0]{--}
\newcommand{\yes}[0]{$\circ$}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 \\
\hline
\hline
\multicolumn{8}{|c|}{clauses} \\
\hline
import w.o prefix & \no & \no & \no & \no & \no & \no & \yes \\
import w prefix & \no & \no & \no & \no & \no & \no & \no \\
variables & \yes & \yes & \yes & \yes & \yes & \yes & \no \\
constants & \no & \yes & \no & \no & \yes & \no & \yes \\
op. input & \yes & \no & \yes & \no & \no & \no & \no \\
op. output & \yes & \no & \yes & \no & \no & \no & \yes \\
\hline
\multicolumn{8}{|c|}{instructions} \\
\hline
block & \yes & \no & \yes & \yes & \no & \yes & \no \\
local var & \no & \no & \yes & \yes & \no & \yes & \yes \\
identity & \no & \no & \yes & \no & \no & \no & \no \\
becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
operation call & \no & \no & \no & \no & \no & \no & \yes \\
condition & \yes & \no & \no & \no & \yes & \no & \yes \\
case & \no & \no & \no & \no & \no & \yes & \no \\
assertion & \no & \no & \no & \no & \no & \no & \no \\
sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes \\
while & \no & \no & \no & \yes & \no & \no & \yes \\
\hline
\multicolumn{8}{|c|}{types} \\
\hline
INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
BOOL & \yes & \yes & \no & \no & \yes & \no & \yes \\
\hline
\multicolumn{8}{|c|}{expressions} \\
\hline
int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes \\
vars & \yes & \yes & \yes & \yes & \no & \yes & \no \\
loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes \\
const & \no & \yes & \no & \no & \yes & \no & \yes \\ 
succ & \no & \yes & \no & \yes & \yes & \no & \no \\
pred & \no & \yes & \no & \yes & \yes & \no & \no \\
+ & \yes & \no & \no & \yes & \no & \yes & \yes \\
- & \yes & \no & \no & \no & \no & \no & \yes \\
* & \no & \no & \no & \no & \no & \yes & \no \\
\hline
\multicolumn{8}{|c|}{formulas} \\
\hline
and & \no & \no & \no & \no & \yes & \no & \no \\
or & \no & \no & \no & \no & \yes & \no & \no \\
not & \no & \no & \no & \no & \yes & \no & \no \\
$<$ & \yes & \yes & \no & \no & \no & \no & \yes \\
$>$ & \no & \no & \no & \no & \yes & \no & \yes \\
\hline
\end{tabular}
\end{center}
